---
title: "EDA"
output:
  pdf_document: default
  html_document: default
date: "2023-09-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

```

```{r}
#Adding random stuff jic we need it
require(tinytex)
require(ggplot2)
require(ggthemes)
require(extrafont)
require(ggstance)
require(fitdistrplus)
require(ggcorrplot)
require(GGally)
require(ggExtra)
require(moments)
require(kable)
require(reshape2)
require(dplyr)
require(corrplot)
require(gridExtra)

```

# Intro

# EDA

words and yk other stuff

```{r}
# Load libraries
library(ggplot2)
library(corrplot)

# Read the data
HDdata <- read.csv("Heart_Disease_Prediction.csv")

# Subset your data
subset_HDdata <- HDdata[, c("Age", "BP", "Cholesterol", "Max.HR", "ST.depression")]

# Create and display histograms, boxplots, and Q-Q plots for all columns
for (col in colnames(subset_HDdata)) {
  # Create histogram with density curve, mean, and standard deviation lines
  histogram_plot <- ggplot(subset_HDdata, aes(x = .data[[col]])) +
    geom_histogram(binwidth = ifelse(col == "ST.depression", 0.5, 5), fill = ifelse(col == "ST.depression", "orange", "blue"), color = "black") +
    geom_density(aes(y = ..density..), kernel = "epanechnikov", size = 1, col = "purple", alpha = 0.2, adjust = 2) +
    geom_vline(xintercept = mean(subset_HDdata[[col]]), col = "red", size = 1, alpha = 0.5) +
    geom_segment(aes(x = mean(subset_HDdata[[col]]) - sd(subset_HDdata[[col]]), y = 0.007, xend = mean(subset_HDdata[[col]]) + sd(subset_HDdata[[col]]), yend = 0.007),
                 col = "red", alpha = 0.3) +
    geom_boxplot(aes(y = -0.005), width = 0.05, notch = TRUE, notchwidth = 0.1, outlier.shape = NA) +
    labs(title = paste("Histogram of", col), x = col, y = "Frequency") +
    theme_minimal()
  
  # Display histogram
  print(histogram_plot)
  
  # Create and display boxplot
  boxplot_plot <- ggplot(subset_HDdata, aes(y = .data[[col]])) +
    geom_boxplot(fill = ifelse(col == "ST.depression", "orange", "blue")) +
    labs(title = paste("Boxplot of", col), x = "", y = col) +
    theme_minimal()
  
  # Display boxplot
  print(boxplot_plot)
  
  # Create and display Q-Q plot
  qq_plot <- ggplot(subset_HDdata, aes(sample = .data[[col]])) +
    geom_qq() +
    labs(title = paste("Normal Q-Q Plot of", col), x = "Theoretical Quantiles", y = "Sample Quantiles") +
    theme_minimal()
  
  # Display Q-Q plot
  print(qq_plot)
}

# Print Summary Statistics
print(summary(subset_HDdata))

# Print Covariance Matrix
print(cov(subset_HDdata))

# Print Correlation Matrix
print(cor(subset_HDdata))

corrplot(cor(subset_HDdata), type = "upper", 
         method = "square", 
         addCoef.col = "white", 
         tl.col = "black", tl.srt = 45)